### 2023-02-23 talk/talk.tex
talk for mfo23

  \begin{enumerate}
  \item When can we carry out full-information likelihood-based inference on a general class of spatiotemporal partially observed Markov process models?

@@

  \item When should we want to?

@@

  \item Is there an intersection between scientific problems of interest and the capabilities of the \texttt{spatPomp} R package?
  \end{enumerate}

  \vspace{10mm}

  We introduce the {\bf iterated block particle filter} because it is currently the most powerful algorithm available in \texttt{spatPomp}.

@@

@@

Question: When should we avoid inference for spatiotemporal models? When do we need to consider coupling? How?

@@

@@

      \item  For time series analysis, iterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do this (\url{https://ionides.github.io/531w22/}). The science may be hard, but the statistics is becoming routine.
      \end{itemize}
  \end{frame}

\begin{frame}{Panel data}

@@

  \item Iterated filtering methods extend to panel data (Breto et al, {\it Journal of the American Statistical Association}, 2019).

@@

  \item Each filter has an iterated version which estimates parameters by repeated filtering using stochastic parameter perturbations.

@@

\begin{frame}{Parameters for the measles model}
  \bi
\item Seasonal tranmission: mean and amplitude, using school term for contact rate.
  \item Durations of latency and infectious period.
\item Initial values: fraction susceptible, latent and infectious.
\item Cohort effect: all births in an age cohort start school in September.
\item Inhomogenous mixing coefficient.
\item Measurement fraction.
\item Population movement model: gravity constant.
\item Dynamic noise (process overdispersion).
\item Measurement overdispersion.

@@

\item BPF was independently proposed as the ``factored particle filter'' by Ng et al (2002, {\it Proc. 18th Conference on Uncertainty and Artificial Intelligence}) but not widely popularized.

@@

@@

    \begin{myitemize}
    \item Blocks in BPF allow recombination (reassortment of chromosomes in sexual reproduction) in the evolutionary analogy.
    \item Blocks are a partition of the metapopulation units. Our experiments suggest treating each sub-population (i.e., city) as a block.

@@

  \item 10-100 coupled units is our target problem size.

@@

@@

  \begin{itemize}
    %$\bullet$
  \item
IBPF can work for simulated data, since it finds likelihoods higher than the truth when the model is well-specified

@@

  \item
    %$\bullet$
    When coupled by a gravity model, simulated data show evidence against a constant immigration model.

@@

  \item 
    %$\bullet$
    Real data do not favor this gravity model over a constant immigration model.
\end{itemize}

@@

%\hspace{-15mm}
\begin{column}{.35\textwidth}
  \vspace{5mm}

$\bullet$  820,000 reported cases and nearly 10,000 deaths

@@

    $\bullet$  Lee et al (2020) developed 3 models to guide vaccination plans:

@@

    {\bf 1}. stochastic, national \\
    {\bf 2}. ODE, spatial \\
    %d\'{e}partment \\
    {\bf 3}. stochastic, spatial
    %d\'{e}partment.

@@

% From Li et al (2020) page S-14:
% According to official report, around 5 million people left Wuhan city before January 23. The total population in Wuhan city after January 23 (when travel restrictions were imposed) is therefore around 6 million.

@@

  \begin{itemize}
  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness, using an iterated EnKF \\
    (Li et al, {\it Science}, May 2020).
   \item
    The time interval predates Wuhan travel restrictions.
%% see README for R code
   \item Total cases 801, Wuhan 454, Chongqing 27, Beijing 26.
   \item{} 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
   \item{} 0 cases reported for all cities during Jan 10--15.

>>>>

  \begin{enumerate}
  \item When can we carry out full-information likelihood-based inference on a general class of spatiotemporal partially observed Markov process models? (Of particular interest are metapopulation models.)

@@

  \item Is there an intersection between scientific problems of interest and the capabilities of the \texttt{spatPomp} R package?

@@

\item We introduce the {\bf iterated block particle filter} because it is currently the most powerful algorithm available in \texttt{spatPomp}.

\item Bonus question: How do we know if our model is statistically adequate, or needs more work?

    \end{enumerate}
\end{frame}

@@

  %% full weekly data now available:
  %% Korevaar, Hannah, C. Jessica Metcalf, and Bryan T. Grenfell. "Structure, space and size: competing drivers of variation in urban and rural measles transmission." Journal of the Royal Society Interface 17, no. 168 (2020): 20200010.
  %% https://royalsocietypublishing.org/doi/pdf/10.1098/rsif.2020.0010
  %% data at https://rs.figshare.com/collections/Supplementary_material_from_Structure_space_and_size_competing_drivers_of_variation_in_urban_and_rural_measles_transmission_/5036567/1

@@

%\vspace{3mm}

%Question: When should we avoid inference for spatiotemporal models? When do we need to consider coupling? How?

@@

      \end{itemize}
  \end{frame}

\begin{frame}{Fitting mechanistic models to time series}

    \begin{itemize}
    \item  Iiterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do plug-and-play likelihood-based inference:

@@

      \url{https://ionides.github.io/531w22/}

      \vspace{5mm}
      \item
      The science may be hard, but the statistics is becoming routine.

    % examples: COVID etc, bison in Yellowstone National Park, volatility in the stock market, gun purchases, human proximal small intestine pH

    \end{itemize}
\end{frame}

\begin{frame}{Fitting mechanistic models to panel data}

@@

  \item Iterated filtering methods extend to panel data (\texttt{panelPomp}, Breto et al, {\it JASA}, 2019).

@@

  \item Iterated filters estimate parameters using stochastic parameter perturbations.

@@

\begin{frame}{Parameters for the measles model}
  \bi
\item Seasonal transmission: mean and amplitude, using school term for contact rate.
  \item Durations of latency and infectious period.
\item Initial values: fraction susceptible, latent and infectious.
\item Cohort effect: all births in an age cohort start school in September.
\item Inhomogenous mixing exponent, $I^{\alpha}$.
\item Measurement fraction.
\item Population movement model: gravity constant.
\item Dynamic noise: process overdispersion.
\item Measurement overdispersion.

\ei
\vspace{3mm}
(Adding gravity coupling to the model of He et al, {\it JRSI}, 2010)
%\vspace{5mm}
%Plug-and-play methods let you add or subtract features

@@

\item BPF was independently proposed as the ``factored particle filter'' by Ng et al (2002).

@@

  \begin{myitemize}
    \item Blocks are a partition of the metapopulation units.
    \item For measles, we use each city as a block.
  \end{myitemize}

@@

    \begin{myitemize}
    \item Blocks are segments of the full state which can be reassorted between particles at the resampling step.

@@

  \item 5-500 coupled units is our target problem size.

@@

%%% A: 9 shared parameters; 4 unit-specific
%%% B: all unit-specific

@@

%%% Wilks' Theorem: improvement on truth is about (1/2) chisq(p)

\begin{itemize}
  \item
IBPF can work for simulated data, since it finds likelihoods higher than the truth when the model is well-specified

@@

  \item
    When coupled by a gravity model, simulated data show evidence against a constant immigration model.

@@

\item 
Data do not favor this gravity model over a constant immigration model.

%%% An underwhelming sounding result, but a necessary step to make progress,
%%% and previously we did not know how to do this.

@@

%\hspace{-15mm}
\begin{column}{.35\textwidth}
  \vspace{2mm}

  $\bullet$  820,000 reported cases

  $\bullet$ nearly 10,000 deaths

@@

    $\bullet$  Lee et al (2020) developed 4 models to guide vaccination plans:

    %%% Elizabeth Lee, Johns Hopkins

@@

    {\bf 1}. stochastic, national \\
    {\bf 2}. ODE, spatial \\
    {\bf 3}. stochastic, spatial\\
    {\bf 4}. agent based

@@

% From Li et al (2020) page S-14:
% Ruiyun Li, Imperial (Oslo as of 2021)
% According to official report, around 5 million people left Wuhan city before January 23. The total population in Wuhan city after January 23 (when travel restrictions were imposed) is therefore around 6 million.

@@

  \begin{itemize}
  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness
    (Li et al, {\it Science}, May 2020).
  \item SEIR model with asymptomatics, reporting delay, and coupling based on cell phone data.
  \item Inference using iterated EnKF.

    \vspace{5mm}
   \item
    The time interval predates Wuhan travel restrictions.
%% see README for R code
   \item Total cases 801, Wuhan 454, Chongqing 27, Beijing 26.
   \item{} 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
   \item{} 0 cases reported for all cities during Jan 10--15.
